---
title: "Exercises 0"
author: "Lúa Arconada Manteca"
date: "2024-04-09"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## EXERCISE 1. Suppose that the joint distribution of a discrete variable, X and two continuous variables Y and Z is

$$
f(x,y,z) \propto \frac{z^{\alpha + x}}{x!}e^{-(\beta + y + 1)z}.
$$

### a) Are X and Y conditionally independent given Z? Why?

We are going to use the criterion we saw in class called 'The factorization theorem'. It says that $X \perp \!\!\! \perp Y |  Z$ if and only if:

$$
f(x,y,z) = g(x,z)h(y,z),
$$
where $g(\cdot)$ and $h(\cdot)$ are not probabilities, but non-negative functions.

Hence, what we need to do, is to factorize the given density function $f(x,y,z)$ as the product of two densities depending on X and Z one and the other on Y and Z.

$$
f(x,y,z) \propto \frac{z^{\alpha + x}}{x!}e^{-(\beta + y + 1)z}.
$$

We can see that we do not have to make any operation since we have the factorization we want. We define:

$$
g(x,z) \propto \frac{z^{\alpha + x}}{x!}
$$

and

$$
h(y,z) \propto e^{-(\beta + y + 1)z}.
$$

The last thing we have to check to be able to use the theorem and conclude that they are conditionally independent is if g and h are non-negative functions. 

Since we said in class that our variables X, Y and Z are non-negative, we have this checked, because $g(x,z)$ is non-negative as a result and so is $h(y,z)$. Hence, X and Y ARE conditionally independent given Z.

### b) Are X and Y independent? Why?

We are going to obtain the joint distribution of X and Y $f_{X,Y}(x,y)$ to see if we can factorize it as the product of the marginals $f_{X,Y}(x,y) = f_X(x)f_Y(y)$, which would indicate independence if it can be done.

We obtain the joint distribution with the following integral:

$$
f_{X,Y}(x,y) = \int_0^{\infty} f(x,y,z) dz    \propto  \int_{0}^{\infty}\frac{z^{\alpha +x}}{x!} e^{-(\beta+y+1)z}dz=\frac{1}{x!}\int_{0}^{\infty}z^{\alpha +x} e^{-(\beta+y+1)z}dz.
$$

To simplify the integral, we make a change of variable setting the new variable as the exponent of the exponential without the minus.

$$
t=(\beta+y+1)z \Rightarrow dt=(\beta+y+1)dz.
$$

Also, we obtain that:

$$
z = \frac{t}{\beta + y + 1}.
$$

Now, we rewrite our integral (without the constant outside of it) with this new variable as follows:

$$
\int_{0}^{\infty}z^{\alpha +x} e^{-(\beta+y+1)z}dz=\int_{0}^{\infty}\frac{t^{\alpha +x}}{(\beta+y+1)^{\alpha+x}} e^{-t}\frac{1}{\beta+y+1}dt=\frac{1}{(\beta+y+1)^{\alpha+x+1}}\int_{0}^{\infty}t^{\alpha +x}e^{-t}dt.
$$

We can see that it is pretty similar to the gamma function, which is as follows:

$$
\Gamma(z)=\int_{0}^{\infty}t^{z-1}e^{-t}dt.
$$

So, we are going to take advantage of this and use it to compute our desired integral, which is solved this way:

$$
\int_{0}^{\infty}z^{\alpha +x} e^{-(\beta+y+1)z}dz=\frac{1}{(\beta+y+1)^{\alpha+x+1}}\int_{0}^{\infty}t^{\alpha +x}e^{-t}dt= \frac{1}{(\beta+y+1)^{\alpha+x+1}}\int_{0}^{\infty}t^{\alpha +x +1 -1}e^{-t}dt= \frac{\Gamma(\alpha+x+1)}{(\beta+y+1)^{\alpha+x+1}}.
$$

To sum all this up, our joint distribution is:

$$
f_{X,Y}(x,y)\propto\frac{\Gamma(\alpha+x+1)}{x!(\beta+y+1)^{\alpha+x+1}},
$$

where we can easily see that it cannot be factorized as our wanted product of the marginal distributions $f_X(x)$ and $f_Y(y)$.

Hence, X and Y ARE NOT independent variables.

### c) Draw an undirected graph to represent the dependence structure of X, Y and Z.

Thanks to parts a) and b), we know that we have to represent the following conditions:

$$
X \perp \!\!\! \perp Y\mid Z\\
X  \not \perp\!\!\!\perp Y
$$

Hence, our corresponding graph is:

```{r 1c}
ug1 <- igraph::graph(c("X","Z", "Y","Z"), directed = FALSE)
plot(ug1)
```

### d) Draw a directed graph to represent the dependence structure of X, Y and Z.

The dependence structure of $X$, $Y$ and $Z$ corresponds to a common parent structure or fork. Hence, our directed acyclic graph (DAG) is the following:

```{r 1d}
dag1 <- gRbase::dag(~ Z*X+Z*Y)
plot(dag1)
```

## EXERCISE 2. Consider the following undirected graph.

```{r 2}
ug1 <- gRbase::ug(c("A","C","D","E"), c("A","B"), c("E","F")) 
ug1 <- gRbase::removeEdge("D", "C", ug1)
ug1 <- gRbase::removeEdge("A", "E", ug1)
ug1 <- gRbase::removeEdge("F", "E", ug1)
plot(ug1)
```

### a) What are the maximal cliques in this graph?

```{r}
iug1 <- as(ug1, "igraph")
igraph::max_cliques(iug1)
```

The maximal cliques are {F}, {B,A}, {A,D}, {A,C}, {C,E} and {D,E}. This is because these are the complete sub graphs to which no node can be added to create a bigger clique.

### b) What is the dependence structure represented by this graph?

To begin with, we can see clearly that B and A are directly dependent, as well as, A and C, C and E, E and D and D and A because they are connected by edges. Moreover, we can see that F is not dependent on any of the other variables. F is completely independent. Furthermore, we can see indirect dependencies between nodes that are connected but not directly by an edge, for example, A and E are indirectly connected.

In addition, we can see than B and D are conditionally independent given A, because all the possible paths between them (B-A-D) go through A. The same with B-C and B-E. The rest of the pairs of variables are not conditionally independent given any other variable because there are always alternative paths.

## EXERCISE 3. Consider the following directed graph.

```{r}
dag1 <- gRbase::dag(~ F+ E:F+ D:E + B+ A:B+ D:A+ C + A:C)
plot(dag1)
```

### a) Are nodes A and F independent? Why?

We check with R.

```{r}
bndag1 <- bnlearn::as.bn(dag1)
bnlearn::dsep(bndag1, "A", "F", character(0))
```

We have obtained that A and F ARE in fact independent, because they are d-separated using an empty set as the conditioning set. Moreover, we can see this because there is no direct path between those two nodes.

### b) Are nodes A and F independent given D? Why?

We check with R.

```{r}
bnlearn::dsep(bndag1, "A", "F", c("D"))
```

A and F are not conditionally independent because they are not d-separated.